preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-14 Manuscript/manuscript.Rmd
updated nodesize
updated text to match fig 5

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was used for model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more frequently co-localized with CTCF sites than either Arrowhead or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more co-localized with other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5D, Supplementary Figure 7). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5E). The better conservation of *preciseTAD*-predicted boundaries further support the notion of their higher biological relevance.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B)Bigwig visualization of known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. (C) Enriched heatmaps comparing the enrichment of CTCF signal around ARROWHEAD TAD boundaries and preciseTAD boundaries, and (D) around Peakachu loop boundaries and preciseTAD for the entire genome. (E) Log 2 genomic distances from boundaries to nearest CTCF sites. (F) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using ARROWHEAD; top left), ARROWHEAD (top right), preciseTAD (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram](Figures/figure5.png)

>>>>

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was used for model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to .01% of the rows in the data. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

Next, we evaluated the biological significance of *preciseTAD* boundaries. Peak regions associated with known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 were found to co-localize more frequently around *preciseTAD* predicted boundaries (Figure 5B). Furthermore, signal heatmaps indicated that our predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Figure 5 C-D; Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5E, Supplementary Figure 7). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5F). The better conservation of *preciseTAD*-predicted boundaries further support the notion of their higher biological relevance.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) Genome browse visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. (C) Enriched heatmaps comparing the enrichment of CTCF signal around ARROWHEAD TAD boundaries and preciseTAD boundaries, and (D) around Peakachu loop boundaries and preciseTAD for the entire genome. (E) Log 2 genomic distances from boundaries to nearest CTCF sites. (F) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using ARROWHEAD; top left), ARROWHEAD (top right), preciseTAD (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram](Figures/figure5.png)
